A global sensitivity analysis of model uncertainty in aeroelastic wind turbine models

### 2020-01-07 Papers/Torque_Paper/article.tex
added discrete routines

Recently, several papers have addressed the uncertainty in BEM model output by performing forward uncertainty propagation or sensitivity studies, e.g.\ \cite{Bos2019a,Echeverria2017,Matthaus2017,Murcia2018,Robertson2018}. In these studies, the focus in mainly on uncertainties in the external conditions (wind parameters) and/or uncertainty in the turbine specification (geometric parameters). Apart from understanding how uncertainty in the output is related to the different uncertain inputs, such sensitivity studies are very useful to reduce the number of parameters as needed for example in design optimization \cite{Echeverria2017}. However, the uncertainty in BEM model output as caused by uncertainty in the model formulation itself, e.g.\ through the values chosen for model parameters, has been given little attention (exceptions being the effect of aerodynamic properties studied in \cite{Matthaus2017,Bortolotti2019}).

@@

\subsection{Geometric uncertainty}
We use Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000} to perturb the geometrical parameters of the turbine blade, such as the reference chord, twist or experimentally obtained drag-, lift-polar curves. The main advantage of using NURBS is that it provides great flexibility to approximate a large variety of curves with a limited number of control points which can be useful in reducing the number of uncertain parameter. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
Use of NURBS has already been exploited in many aerodynamics applications, for instance, design optimization of wind-turbine blades, see for e.g. \cite{Ribeiro2012,Bottasso2014} and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize curves using NURBS.
\subsubsection{NURBS based perturbation}
The value of NURBS curve at each location $x$ is computed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(x)  = \sum_{i=1}^{n} c_i B_{i,p}(x),
\end{equation}
where $S(x)$ is the value of the NURBS curve at location $x$; $c_i$ is the weight of the $i$-th control point and $B_{i,p}(x)$ is the value of B-spline corresponding to the $i$-th control point at $x$ with the subscript $p$ denoting its polynomial degree. A sequence of non-decreasing parameters $X = \{x_1,x_2,...,x_m\}$ called the \textit{knot vector} determines the domain in which any control point is active with the number of elements in the knot vector is given by $m = n+p+1$. B-splines are non-zero only within a knot interval $[x_i,x_{i+1})$ for $i = 1, 2, ..., m-1$ and are zero elsewhere. Therefore, only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}. For a curve of polynomial degree p, there are $p+1$ active control points at any location $x$, for e.g., for a linear curve ($p=1$) only two adjacent control points are active at any location. Also, note that the number of B-splines are equal to the number of elements in $X$. Another useful property of the B-splines is that they are recursive in polynomial degree thus we can derive quadratic B-splines using linear B-splines, cubic from quadratic B-splines and so on. Starting with B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(x) :=
\begin{cases}
1\quad x_i\leq x < x_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
we can derive higher order B-splines in the knot interval $[x_{i+p},x_{i+p+1})$ using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(x) := \frac{x - x_i}{x_{i+p} - x_i}B_{i,p-1}(x) + \frac{x_{i+p+1}  -  x}{x_{i+p+1}  -  x_{i+1}}B_{i+1,p-1}(x),\quad p\geq1.
\end{equation} 
These B-splines can be constructed efficiently using the de Boor's algorithm \cite{deBoor}. Note that to perform above computation, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. In Fig. \ref{B-splines}, we show linear, quadratic and cubic splines for $x\in[0,1]$ [Also show the padding].

@@

\subsection{Model uncertainty}
\subsection{Wind velocity}
\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or in combination) and how variations in input values affects the output of interest. In engineering, sensitivity analysis can be employed for a number of reasons: to determine the stability and robustness of a computational model with respect to input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters. Sensitivity analysis techniques can be classified as local and global methods, see \cite{RSmith}. In local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighbourhood of nominal input values. Although, local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify global behavior of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, global sensitivity analysis is more suitable for aeroelastic models considered in this work.

@@

\subsubsection{Ordinary Least Square (PCE\_OLS)}
\subsubsection{Least Angle Regression (PCE\_LAR)}
%\section{Description of test case}
%\section{Sensitivity analysis workflow}
\section{Results}\label{sec:results}

>>>>

Recently, several papers have addressed the uncertainty in BEM model output by performing forward uncertainty propagation or sensitivity studies, e.g.\ \cite{Echeverria2017,Matthaus2017,Murcia2018,Robertson2018,Bos2019a}. In these studies, the focus in mainly on uncertainties in the external conditions (wind parameters) and/or uncertainty in the turbine specification (geometric parameters). Apart from understanding how uncertainty in the output is related to the different uncertain inputs, such sensitivity studies are very useful to reduce the number of parameters as needed for example in design optimization \cite{Echeverria2017}. However, the uncertainty in BEM model output as caused by uncertainty in the model formulation itself, e.g.\ through the values chosen for model parameters, has been given little attention (exceptions being the effect of aerodynamic properties studied in \cite{Matthaus2017,Bortolotti2019}).

@@

\subsection{Geometric uncertainty}
We use Non-Uniform Rational Basis Splines (NURBS) \cite{rogers2000} to perturb the geometrical parameters of the turbine blade, such as the reference chord, twist or experimentally obtained drag-, lift-polar curves. The main advantage of using NURBS is that it provides great flexibility to approximate a large variety of curves with a limited number of control points which can be useful in reducing the number of uncertain parameter. %Further, the set of control points and knots can be directly manipulated to control the smoothness and curvature.
Use of NURBS has already been exploited in many aerodynamics applications, for instance, design optimization of wind-turbine blades, see for e.g. \cite{Ribeiro2012,Bottasso2014} and to parametrize chord and twist curves \cite{Echeverria2017}. In the following, we briefly outline the procedure used to parametrize curves using NURBS.
\subsubsection{NURBS based perturbation}
The value of NURBS curve at each location $x$ is computed using a weighted sum of $n$ basis functions (or B-splines):
\begin{equation}\label{NURB_curve}
S(x)  = \sum_{i=1}^{n} c_i B_{i,p}(x),
\end{equation}
where $S(x)$ is the value of the NURBS curve at location $x$; $c_i$ is the weight of the $i$-th control point and $B_{i,p}(x)$ is the value of B-spline corresponding to the $i$-th control point at $x$ with the subscript $p$ denoting its polynomial degree. A sequence of non-decreasing parameters $X = \{x_1,x_2,...,x_m\}$ called the \textit{knot vector} determines the domain in which any control point is active with the number of elements in the knot vector is given by $m = n+p+1$. B-splines are non-zero only within a knot interval $[x_i,x_{i+1})$ for $i = 1, 2, ..., m-1$ and are zero elsewhere. Therefore, only a few nearby control points are active when computing the NURBS curve \eqref{NURB_curve}. For a curve of polynomial degree $p$, there are $p+1$ active control points at any location $x$, for e.g., for a linear curve ($p=1$) only two adjacent control points are active at any location. Also, note that the number of B-splines are equal to the number of elements in $X$. Another useful property of the B-splines is that they are recursive in polynomial degree thus we can derive quadratic B-splines using linear B-splines, cubic from quadratic B-splines and so on. Starting with B-spline of degree 0 defined as 
\begin{equation}\label{linearBspline}
B_{i,0}(x) :=
\begin{cases}
1\quad x_i\leq x < x_{i+1},\quad i = 1,2,...,m,\\
0\quad\text{elsewhere,}
\end{cases} 
\end{equation}
we can derive higher order B-splines in the knot interval $[x_{i+p},x_{i+p+1})$ using the recurrence relation \cite{deBoor}:
\begin{equation}\label{NURBS_recurrence}
B_{i,p}(x) := \frac{x - x_i}{x_{i+p} - x_i}B_{i,p-1}(x) + \frac{x_{i+p+1}  -  x}{x_{i+p+1}  -  x_{i+1}}B_{i+1,p-1}(x),\quad p\geq1.
\end{equation} 
These B-splines can be constructed efficiently using the de Boor's algorithm \cite{deBoor}. Note that to perform above computation, we need to pad the knot vector $X$ by repeating the first and last elements $p$ times each. In Fig. \ref{B-splines}, we show linear, quadratic and cubic splines for $x\in[0,1]$ [Also show the padding].

@@

There are two ways to obtain perturbed curves either by independently perturbing each of the control points (local perturbation) or using the same random number to perturb all the control points (global perturbation). We use local perturbation for chord control points Ch1-Ch5 and twist control points Tw1-Tw6 control points, see Fig. \ref{perturbed_samples} (a) - (b), respectively. Each control points are perturbed using a uniform random number with bounds given by $\pm5\%$ of the magnitude of control point. For lift and drag curve, all control points are perturbed using same uniform random number resulting in samples of the lift and drag curves as shown in Fig. \ref{perturbed_samples} (c) - (d), respectively. For simplicity, we only consider uniform distributions for perturbing the control points, however, we can also use other probability distributions for this. Also note that when using a uniform distribution with bounds given by relative magnitude of control points, we tend to obtain small perturbations for control points with values close to zero.
\begin{figure}[h!]
\centering
\includegraphics[trim={0 1.6cm 0 1.5cm},clip, scale=0.67]{figure1.pdf}
\caption{Random realization of chord, twist, lift- (Cl2) and drag-polars (Cd2).}
\label{perturbed_samples}
\end{figure}

\section{Global sensitivity analysis}\label{sec:GSA}
The objective of sensitivity analysis is to quantify the relative significance of individual inputs (or in combination) and how variations in input values affects the output of interest. In engineering, sensitivity analysis can be employed for a number of reasons: to determine the stability and robustness of a computational model with respect to input parameters, for simplification of stochastic models by fixing the insensitive parameters, and to guide data acquisition campaigns and experimental design to refine the data on sensitive parameters. Sensitivity analysis techniques can be classified as local and global methods, see \cite{RSmith}. In local sensitivity analysis, individual parameters are perturbed around their nominal values allowing for the description of output variability only in a small neighbourhood of nominal input values. Although, local approaches are widely employed due to their ease of implementation and low computational cost, they are unable to quantify global behavior of nonlinearly parametrized models such as aeroelastic models. Global sensitivity approaches, on the other hand, consider the entire range of input values to compute output sensitivities. Therefore, global sensitivity analysis is more suitable for aeroelastic models considered in this work.

@@

Quadrature based integration methods suffer from curse of dimensionality where the number of integration points rises exponentially with an increase in the number of dimensions. For applications with a fixed computational budget methods that minimizes the approximation error with a given number of samples are usually more practical. Next, we briefly describe two such methods to compute PCE coefficients based on a given number of model evaluations.

\subsubsection{Ordinary Least Square}
The 
\subsubsection{Least Angle Regression (PCE\_LAR)}
\subsection{Sensitivity analysis with discrete random variable}
%\section{Description of test case}
%\section{Sensitivity analysis workflow}
\section{Results}\label{sec:results}
